### 2024-04-14 sections/4-results.tex
paper edits 4.3-end

Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for (the non-augmented versions of) Fashion-Mnist, Glove-25, Sift, Random, Silva 18S, and RadioML.
The horizontal axis denotes the depth in the cluster tree, and the vertical axis denotes the LFD (as calculated by Equation~\ref{eq:methods:lfd-simplified}) of clusters at that depth.
We plot lines for the 5$^{th}$, 25$^{th}$, 50th, 75$^{th}$ and 95$^{th}$ percentiles of LFD, as well as the minimum and maximum LFD at each depth.
In order to have the plots to best reflect the distribution of LFDs across the entire \textit{dataset}, we compute percentiles counting each cluster as many times as its cardinality.
In other words, if, for some dataset, the 95$^{th}$ percentile of LFD at depth 40 is 3, this means that 95\% of the points in clusters at depth 40 belong to clusters whose LFD is at most 3.

@@

In contrast with Figure~\ref{fig:results:sift-lfd}, Figure~\ref{fig:results:random-lfd} shows the LFD by depth for a random dataset with the same cardinality and dimensionality as those of Sift.
We observe that the LFD starts as high as 20 (for all data) at depth 0.
As shown by the needlepoint shape of the plot, the LFD for all clusters has a very narrow spread from depth 0 through depth 5.
After depth 5, we begin to see a wider spread between the maximum and minimum LFD at each depth.
All percentile lines seem to decrease linearly with depth.
The LFD of approximately 20 at depth 0 (i.e. the root cluster) is what we expect for this random dataset.
To elaborate, the distribution of points in such a dataset should reflect the curse of dimensionality, i.e. the fact that in high dimensional spaces, the minimum and maximum pairwise distances between any two points are approximately equal.
As a result, the root cluster's radius $r$, which reflects the maximum distance between the center $c$ and any other point, should not differ significantly from the distance between the center and its closest point.
A consequence of this is that, with high probability, for every point in the root cluster, its distance from $c$ is greater $\tfrac{r}{2}$;
$B_X(c, \tfrac{r}{2})$ contains only $c$ while $B_X(c, r)$ contains the entire dataset.
Given our definition of LFD in Equation~\ref{eq:methods:lfd-original}, this means that the LFD of the root cluster is approximately $\log_2(\frac{|X|}{1}) = \log_2(1,000,000) \approx 20$, which is what we observe in Figure~\ref{fig:results:random-lfd}.

@@

The horizontal axis in each figure shows the cardinality of the augmented dataset with synthetic points (see Section \ref{sec:methods:synthetic-data}).
The left-most point on each line is at the cardinality of the original dataset without any synthetic augmentation.
The vertical axis denotes throughput in queries per second.
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k = 100$ can be found in the Supplement.
For HNSW and ANNOY, we report the recall for each measurement in the plots. For algorithms which exhibit recall greater than $0.9995$, we do not report the recall in the plots, but we do report it in the tables below.

@@

Figure~\ref{fig:results:scaling-plots} shows that for Fashion-Mnist, Glove-25, and Sift, as cardinality increases, the CAKES algorithms (Depth-First Sieve in blue, Repeated $\rho$-NN in green, and Breadth-First Sieve in purple) become faster than our Rust implementation of na\"{i}ve linear search (in orange).
Though we observe this trend, we note that the exact cardinality at which CAKES's algorithms overtake linear search differs by dataset. For Fashion-Mnist, CAKES begins exhibiting sub-linear performance starting at a cardinality near $10^5$, while for Glove-25 and Sift, this happens near $10^6$ and $10^7$ respectively. 
Which one of the three CAKES algorithms is fastest also differs by dataset.
For Fashion-Mnist, Depth-First Sieve is consistently fastest, while for Glove-25, the fastest is Repeated $\rho$-NN, and for Sift, Breadth-First Sieve. 
Additionally, on all three datasets, Depth-First Sieve and Breadth-First Sieve appear to have performance which is \textit{constant} in the cardinality of the dataset. 
With Glove-25, Repeated $\rho$-NN exhibits similar constant scaling. 
We also observe that on these three datasets, for nearly all cardinalities, all three of the CAKES algorithms are faster than FAISS-Flat (in brown), and that at some cardinality, CAKES's algorithms become faster than FAISS-IVF (in pink). 
For Fashion-Mnist, CAKES becomes faster than FAISS-IVF near cardinality $10^5$, whereas for Glove-25 and Sift, this happens near cardinality $10^7$. 
On all three of these datasets, HNSW (in gray) and ANNOY (in yellow) are faster than CAKES's algorithms for all cardinalities; however, CAKES exhibits perfect or near-perfect recall on each dataset, while HNSW and ANNOY exhibit much lower recall, as shown in Table~\ref{tab:datasets:summary}.
While recall for CAKES's algorithms does not degrade with cardinality, recall for HNSW and ANNOY does degrade with cardinality. 
At a cardinality multiplier as low as eight, HNSW and ANNOY have recall of $0.525$ and $0.857$ respectively for Fashion-Mnist, 0.607 and 0.832 for Glove-25, and 0.782 and 0.686 on Sift. 
In contrast, CAKES has perfect recall on Fashion-Mnist and Sift (because the distance function used with these datasets is a metric), and near-perfect recall on Glove-25 (because the Cosine distance function is not a metric).

@@

For Silva and RadioML, we benchmarked only CAKES's algorithms because HNSW, ANNOY and FAISS do not support neither the required distance functions nor, in the case of RadioML, complex-valued data.
Due to the massive sizes of these datasets, we took random sub-samples of the dataset with lower cardinalities for our benchmarks, rather than augmented versions of the dataset.
With Silva, as shown in Figure~\ref{fig:results:silva-scaling}, we observe that for all algorithms, throughput seems to linearly decrease as cardinality increases, but that it seems to begin levelling off near cardinality $10^5$.
Until cardinality near $10^4$, Depth-First Sieve is the fastest CAKES algorithm, but for larger cardinalities, Repeated $\rho$-NN is the fastest CAKES algorithm.
For RadioML, as shown in Figure~\ref{fig:results:radioml-scaling}, we observe that throughput declines nearly linearly the three CAKES algorithms exhibit nearly indistinguishable performance for all cardinalities.
In particular, throughput of CAKES's algorithms is identical within three significant figures for all cardinalities.

>>>>

Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for (the non-augmented versions of) Fashion-Mnist, Glove-25, Sift, Random, Silva 18S, and RadioML.
The horizontal axis denotes the depth in the cluster tree, and the vertical axis denotes the LFD (as calculated by Equation~\ref{eq:methods:lfd-simplified}) of clusters at that depth.
We plot lines for the 5$^{th}$, 25$^{th}$, 50th, 75$^{th}$ and 95$^{th}$ percentiles of LFD, as well as the minimum and maximum LFD at each depth.
In order to have the plots best reflect the distribution of LFDs across the entire \textit{dataset}, in computing percentiles, we count each cluster as many times as its cardinality.
In other words, if, for some dataset, the 95$^{th}$ percentile of LFD at depth 40 is 3, this means that 95\% of the points in clusters at depth 40 belong to a cluster whose LFD is at most 3.

@@

In contrast with Figure~\ref{fig:results:sift-lfd}, Figure~\ref{fig:results:random-lfd} shows the LFD by depth for a random dataset with the same cardinality and dimensionality as those of Sift.
We observe that the LFD starts as high as 20 (for all data) at depth 0.
As shown by the needlepoint shape of the plot, the LFD for all clusters has a very narrow spread from depth 0 through depth 5.
After depth 5, we begin to see a wider spread between the maximum and minimum LFD at each depth.
All percentile lines seem to decrease linearly with depth.
The LFD of approximately 20 at depth 0 (i.e. the root cluster) is what we expect for this random dataset.
To elaborate, the distribution of points in such a dataset should reflect the curse of dimensionality, i.e. the fact that in high dimensional spaces, the minimum and maximum pairwise distances between any two points are approximately equal.
As a result, the root cluster's radius $r$, which reflects the maximum distance between the center $c$ and any other point, should not differ significantly from the distance between the center and its closest point.
A consequence of this is that, with high probability, for every point in the root cluster, its distance from $c$ is greater than $\tfrac{r}{2}$;
in other words, $B_X(c, \tfrac{r}{2})$ contains only $c$ while $B_X(c, r)$ contains the entire dataset.
Given our definition of LFD in Equation~\ref{eq:methods:lfd-original}, this means that the LFD of the root cluster is approximately $\log_2(\frac{|X|}{1}) = \log_2(1,000,000) \approx 20$, which is what we observe in Figure~\ref{fig:results:random-lfd}.

@@

The horizontal axis in each figure shows the cardinality of the dataset augmented with synthetic points (see Section \ref{sec:methods:synthetic-data}).
The left-most point on each line is at the cardinality of the original dataset without any synthetic augmentation.
The vertical axis denotes throughput in queries per second.
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k = 100$ can be found in the Supplement.
For HNSW and ANNOY, we report the recall for each measurement in the plots. For algorithms which exhibit recall greater than $0.9995$, we do not report the recall in the plots, but we do report it in the tables below.

@@

Figure~\ref{fig:results:scaling-plots} shows that for Fashion-Mnist, Glove-25, and Sift, as cardinality increases, the CAKES algorithms (Depth-First Sieve in blue, Repeated $\rho$-NN in green, and Breadth-First Sieve in purple) become faster than our Rust implementation of na\"{i}ve linear search (in orange).
Though we observe this trend, we note that the exact cardinality at which CAKES's algorithms overtake linear search differs by dataset. For Fashion-Mnist, CAKES begins exhibiting sub-linear performance starting at a cardinality near $10^5$, while for Glove-25 and Sift, this happens near $10^6$ and $10^7$ respectively. 
Which one of the three CAKES algorithms is fastest also differs by dataset.
For Fashion-Mnist, Depth-First Sieve is consistently fastest, while for Glove-25, the fastest is Repeated $\rho$-NN, and for Sift, Breadth-First Sieve. 
Additionally, on all three datasets, Depth-First Sieve and Breadth-First Sieve appear to have performance which is \textit{constant} in the cardinality of the dataset. 
With Glove-25, Repeated $\rho$-NN exhibits similar constant scaling. 
We also observe that on these three datasets, for nearly all cardinalities, all three of the CAKES algorithms are faster than FAISS-Flat (in brown), and that at some cardinality, CAKES's algorithms become faster than FAISS-IVF (in pink). 
For Fashion-Mnist, CAKES becomes faster than FAISS-IVF near cardinality $10^5$, whereas for Glove-25 and Sift, this happens near cardinality $10^7$. 
On all three of these datasets, HNSW (in gray) and ANNOY (in yellow) are faster than CAKES's algorithms for all cardinalities; however, CAKES exhibits perfect or near-perfect recall on each dataset, while HNSW and ANNOY exhibit much lower recall, as shown in Table~\ref{tab:datasets:summary}.
While recall for CAKES's does \emph{not} degrade with cardinality, recall for HNSW and ANNOY does degrade with cardinality. 
At a cardinality multiplier as low as eight, HNSW and ANNOY have recall of $0.525$ and $0.857$ respectively for Fashion-Mnist, 0.607 and 0.832 for Glove-25, and 0.782 and 0.686 on Sift. 
In contrast, CAKES has perfect recall on Fashion-Mnist and Sift (because the distance function used with these datasets is a metric), and near-perfect recall on Glove-25 (because the Cosine distance function is not a metric).

@@

For Silva and RadioML, we benchmarked only CAKES's algorithms because HNSW, ANNOY and FAISS do not support neither the required distance functions nor, in the case of RadioML, complex-valued data.
Due to the massive sizes of these datasets, we took random sub-samples of the dataset--rather than augmented versions of the dataset--to examine how performance scales with cardinality.
With Silva, as shown in Figure~\ref{fig:results:silva-scaling}, we observe that for all algorithms, throughput seems to linearly decrease as cardinality increases, but that it seems to begin levelling off near cardinality $10^5$.
Until cardinality near $10^4$, Depth-First Sieve is the fastest CAKES algorithm, but for larger cardinalities, Repeated $\rho$-NN is the fastest CAKES algorithm.
For RadioML, as shown in Figure~\ref{fig:results:radioml-scaling}, we observe that throughput declines nearly linearly, and that the three CAKES algorithms exhibit virtually indistinguishable performance for all cardinalities.
In particular, throughput of CAKES's algorithms is identical within three significant figures for all cardinalities.

### 2024-04-14 sections/5-discussion.tex
paper edits 4.3-end

CAKES's algorithms are designed to be most effective when the data uphold the manifold hypothesis, or in other words, when the data are constrained to a low-dimensional manifold even when embedded in a high-dimensional space.
As a consequence, these algorithms do not perform well on data with random distributions, because of the absence of a manifold structure.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each of the datasets we tested exhibits a manifold structure, as quantified by the percentiles of LFD.
As expected, our results from Figure~\ref{fig:results:scaling-plots} indicate that CAKES's algorithms scale sub-linearly with cardinality on datasets with low LFD, and scale linearly on datasets with high LFD.

As seen in Figures~\ref{fig:results:random-lfd}~and~\ref{fig:results:sift-lfd}, the LFD of the Random dataset is much higher than that of Sift, even though they have the same cardinality and embedding dimension. 
This is unsurprising, given that the Random dataset is, by definition, uniformly distributed, while Sift is not.
On the Random dataset, CAKES's algorithms' speed decreases linearly as the multiplier increases, while with Sift, Depth-First Sieve and Breadth-First Sieve both exhibit nearly constant throughput.
While Repeated $\rho$-NN does not exhibit near-constant throughput with Sift, the throughput decreases much more slowly than it does with the Random dataset.
Since these two datasets have the same cardinality and embedding dimension, the aforementioned discrepancies show the effect of a manifold structure on algorithm performance.
We emphasize that even though CAKES's algorithms are slower on the Random dataset, their recalls remain perfect at all multipliers.

@@

When written with the same notation as used in Section~\ref{sec:methods:knn-search:complexity-of-sieve-methods}, we see that the time complexity of Repeated $\rho$-NN is $\mathcal{O}(\tfrac{\mathcal{T}}{d} + \mathcal{L})$, where $\mathcal{T}$ is the time complexity of tree-search and $\mathcal{L}$ is the time complexity of leaf-search. 
Even though Repeated $\rho$-NN has the lowest time complexity (compare to $\mathcal{O}(\mathcal{T}\log{(\mathcal{T}+\mathcal{L})} + \mathcal{L}\log{(\mathcal{T}+\mathcal{L})})$ for Breadth-First Sieve and $\mathcal{O}(\mathcal{T}\log{\mathcal{T}} + \mathcal{L}\log{k})$ for Depth-First Sieve) of the three CAKES algorithms, it is not always the fastest algorithm empirically. 
We believe that some of this discrepancy can be explained by the fact that if Repeated $\rho$-NN significantly ``overshoots'' the correct radius for $k$ hits ($\rho_k$) during tree-search, leaf-search will require looking at more points, rendering the true scaling factor higher than that in~\ref{eq:methods:repeated-rnn-complexity}.
This overshot can occur when the LFDs of clusters near the query are not concentrated around their expectation. 
For example, if most clusters near the query have very low LFD except for one anomaly with very high LFD, 
the harmonic mean LFD $\mu$ can still be low, so the factor of radial increase~\ref{eq:methods:repeated-rnn-factor} may be much larger than necessary for guaranteeing $k$ hits.  
This suggests that rather than use the reciprocal of the \textit{harmonic} mean LFD in~\ref{eq:methods:repeated-rnn-factor}, we may actually want a mean which is more sensitive to high outliers, such as the geometric mean.
We leave it as an avenue for future work to characterize when Repeated $\rho$-NN significantly ``overshoots'' $\rho_k$ and to improve upon the factor of radial increase in~\ref{eq:methods:repeated-rnn-factor} so that this overshooting occurs less frequently and with less severity.

@@

We also plan to investigate hierarchical data compression by representing differences at each level of the binary tree, particularly for string or genomic data, where all differences are discrete.
Such an encoding would allow us to store the data in a compressed format, and to perform search on the compressed data without decompressing it.
This would allow us to perform fast search on datasets that are too large to fit in memory.
An exploration of compression and compressed search is an avenue for future work.

Another avenue for future work is to explore the use of CAKES in a streaming environment.
This would require the ability to perform ``online'' updates to the tree as points are added to or deleted from the dataset.
Such online updates would take advantage of the fast search algorithms provided by CAKES.  % and would allow us to have our cake and eat it, too.
CAKES can also be used to extend anomaly detection in CHAODA~\cite{ishaq2021clustered}.
We could add to CHAODA's ensemble of graph-based anomaly detection methods by using the distribution of distances among the $k$ nearest neighbors of cluster centers.

>>>>

CAKES's algorithms are designed to be most effective when the data uphold the manifold hypothesis, or in other words, when the data are constrained to a low-dimensional manifold even when embedded in a high-dimensional space.
As a consequence, these algorithms do not perform well on data with random distributions, because of the absence of a manifold structure.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.
As expected, our results from Figure~\ref{fig:results:scaling-plots} indicate that CAKES's algorithms scale sub-linearly with cardinality on datasets with low LFD, and scale linearly on datasets with high LFD.

As seen in Figures~\ref{fig:results:random-lfd}~and~\ref{fig:results:sift-lfd}, the LFD of the Random dataset is much higher than that of Sift, even though both datasets have the same cardinality and embedding dimension. 
This is unsurprising, given that the Random dataset is, by definition, uniformly distributed, while Sift is not.
On the Random dataset, CAKES's algorithms' speed decreases linearly as the multiplier increases, while with Sift, Depth-First Sieve and Breadth-First Sieve both exhibit nearly constant throughput.
While Repeated $\rho$-NN does not exhibit near-constant throughput with Sift, the throughput decreases much more slowly than it does with the Random dataset.
Since these two datasets have the same cardinality and embedding dimension, the aforementioned discrepancies highlight how manifold structure affects algorithm performance.
We emphasize that even though CAKES's algorithms are slower on the Random dataset, their recalls remain perfect at all multipliers.

@@

When written with the same notation as used in Section~\ref{sec:methods:knn-search:complexity-of-sieve-methods}, we see that the time complexity of Repeated $\rho$-NN is $\mathcal{O}(\tfrac{\mathcal{T}}{d} + \mathcal{L})$, where $\mathcal{T}$ is the time complexity of tree-search and $\mathcal{L}$ is the time complexity of leaf-search. 
Even though Repeated $\rho$-NN has the lowest time complexity (compare to $\mathcal{O}(\mathcal{T}\log{(\mathcal{T}+\mathcal{L})} + \mathcal{L}\log{(\mathcal{T}+\mathcal{L})})$ for Breadth-First Sieve and $\mathcal{O}(\mathcal{T}\log{\mathcal{T}} + \mathcal{L}\log{k})$ for Depth-First Sieve) of the three CAKES algorithms, it is not always the fastest algorithm empirically. 
We believe that some of this discrepancy can be explained by the fact that if Repeated $\rho$-NN significantly ``overshoots'' the correct radius for $k$ hits ($\rho_k$) during tree-search, leaf-search will require looking at more points, rendering the true scaling factor higher than that in~\ref{eq:methods:repeated-rnn-complexity}.
This overshot can occur when the LFDs of clusters near the query are not concentrated around their expectation. 
For example, if most clusters near the query have very low LFD except for one anomaly with very high LFD, 
the harmonic mean LFD $\mu$ can still be low, so the factor of radial increase~\ref{eq:methods:repeated-rnn-factor} may be much larger than necessary for guaranteeing $k$ hits.  
This suggests that rather than using the reciprocal of the \textit{harmonic} mean LFD in~\ref{eq:methods:repeated-rnn-factor}, we may achieve better results with a mean that is more sensitive to high outliers, such as the geometric mean.
We leave it as an avenue for future work to characterize when Repeated $\rho$-NN significantly ``overshoots'' $\rho_k$ and to improve upon the factor of radial increase in~\ref{eq:methods:repeated-rnn-factor} so that this occurs less frequently and with less severity.

@@

We also plan to investigate hierarchical data compression by representing differences at each level of the binary tree, particularly for string or genomic data, where all differences are discrete.
Such an encoding would allow us to store the data in a compressed format, and to perform search on the compressed data without decompressing it.
This would allow us to perform fast search on datasets that are too large to fit in memory.
An exploration of compression and compressed search is another avenue for future work.

We also would like to explore the use of CAKES in a streaming environment.
This would require the ability to perform ``online'' updates to the tree as points are added to or deleted from the dataset.
Such online updates would take advantage of the fast search algorithms provided by CAKES.  % and would allow us to have our cake and eat it, too.
CAKES can also be used to extend anomaly detection in CHAODA~\cite{ishaq2021clustered}.
We could add to CHAODA's ensemble of graph-based anomaly detection methods by using the distribution of distances among the $k$ nearest neighbors of cluster centers.
